Manuscript draft

### 2024-02-28 manuscript.qmd
added some figures after sourcing modeling code

```{r}
#| include: false
library(tidyverse)
```

@@

Social situations occur over different time scales, and this is an important consideration when studying the causes and consequences of their structure. In defining co-roosting, co-flight, and co-feeding interactions based on proximity in each of these behavioral situations, we have created social networks in which edges already conceptually represent interactions over different amounts of time. Because individuals have one roosting location per night, edge weights in the co-roosting network are measured in units of days, while pairs of individuals can interact many times in flight or at a carcass on a given day.\
\
The biological processes driving the execution of these behaviors and the formation of these networks are likewise happening over different time scales. An individual must decide once per day which roost to go to [@harel2016]. But flight is a continuous process driven by many decisions resulting from individual and social information [@sassi2023], and feeding events are concentrated in time on a given day. Feeding events are driven by the temporal rhythm of the appearance of carcasses, which occur naturally or are provisioned every [NUMBER] days and persist for hours to a few days (Orr Spiegel, pers. comm.). Individuals may return to the same carcass on successive days. While at a carcass, individuals may fight and interact many times.

@@

To answer these questions, we drew on data from a free-flying population of Eurasian griffon vultures (Gyps fulvus) fitted with GPS tags. \[tag specs\]. Over the period 2020-09-01 to 2023-05-15, the dataset contained \[number\] of tagged vultures. Tags recorded GPS fixes, as well as speed and heading readings, every 10 minutes when battery conditions allowed. The GPS data are stored on Movebank [@kays2022] and were loaded into R using the `move` package\[ @kranstauber2023\].

@@

We subsetted the dataset to exclude periods of time that vultures spent in capture cages, in the wildlife hospital following poisonings or injuries, or in captivity prior to wild release (in the case of captive-born birds). We cleaned the remaining data to remove invalid GPS fixes \[more details\]. Then we split data from each into three seasons. The fall, or "pre-breeding," season includes dates from 15 September through 14 December; the breeding season lasts from 15 December through 14 May of the following year; and the summer lasts from 15 May through 14 September. These season boundaries were chosen based on biological knowledge of the vultures' breeding timelines (see more details in the Study System section). Therefore, the analysis contained a total of 8 seasons, from the fall of 2020 through the 2023 breeding season (Table X). All of these data was reserved for use in proximity-based social network analysis. Since the GPS tags give few to no fixes overnight, and since reception is often very poor on the cliffs where vultures roost, we combined information from individuals' last locations of the evening and first locations of the morning to assign a roost site for each individual on each night \[more details\]. Full details about the data cleaning and roost assignment procedure can be found in the supplementary material.

@@

For each season, we calculated movement measures only for individuals tracked for at least 30 days (after removing low-point days as described above). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the `elevatr` R package [@hollister2022], at a zoom level of [level–more details about resolution?] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

@@

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. Using the `adehabitatHR` package [@calenge2023], we calculated home range as the 95% kernel utilization distribution (KUD) of a vulture's GPS fixes in a season; core area as the 50% KUD, and core area use intensity as the ratio of the 50% to the 95% KUD.

@@

To gain a complete picture of individual vultures' spatial phenotype, we summarized the above-described movement measures (centered and scaled) in a principal component analysis. The first two principal components together explained 65% of the variance in spatial phenotype (Table X).

PC1 explained 47.6% of the variance in spatial phenotype. All movement variables except tortuosity loaded positively on PC1, with the strongest contributors being mean daily maximum displacement, roost diversity, and proportion of nights roost-switching (Figure X). Overall, we can understand PC1 as an aggregate measure of individual movement. Vultures with high PC1 values cover more area, use more different roosts, travel farther, spend more time flying, and have straighter trajectories.

PC2 explained 17.4% of the variance in spatial phenotype. The space use variables loaded positively on this principal component, while flight variables (both altitude measures, daily distance traveled, and daily flight time) contributed strongly negatively (Figure X). We can understand PC2 as a rough measure of exploratory behavior. Vultures with high PC2 values use a larger spatial area, while those with low PC2 values fly higher and spend more time flying (in a more concentrated area).

@@

```{r}
#| echo: false
#| output: true
#| label: fig-nvultures
#| fig-cap: "Number of vultures in different social situations over the study period. Lines for the 'feeding' situation have been jittered upward because they overlap completely with the 'flight' situation. Dashed lines and open circles represent the total number of vultures tracked in the population. Solid lines and filled circles represent the number of focal vultures."

load("data/derived/forModeling.Rda")
load("data/derived/cc.Rda")
load("data/derived/seasonNames_orig.Rda")
seasonNames <- seasonNames_orig[-1]
focals <- forModeling %>%
  filter(!is.na(PC1)) %>%
  group_by(season, year, type) %>%
  summarize(total = n[1],
            focal = length(unique(Nili_id)), .groups = "drop") %>%
  arrange(year, season, type) %>%
  mutate(seasonUnique = paste(year, season, sep = "_")) %>%
  mutate(seasonUnique = factor(seasonUnique, levels = c("2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall", "2023_breeding"))) %>%
  mutate(seasonU = str_replace(seasonUnique, "breeding", "B")) %>%
  mutate(seasonU = str_replace(seasonU, "fall", "F")) %>%
  mutate(seasonU = str_replace(seasonU, "summer", "S")) %>%
  mutate(seasonU = str_remove(seasonU, "_")) %>%
  mutate(seasonU = str_remove(seasonU, "(?<=^)20")) %>%
  mutate(seasonU = factor(seasonU, levels = c("20F", "21B", "21S", "21F", "22B", "22S", "22F", "23B")))  %>%
  pivot_longer(cols = c(total, focal), names_to = "Vultures", values_to = "n") %>%
  rename("Situation" = "type") %>%
  ungroup()

focals %>%
  mutate(n = case_when(Situation == "feeding" ~ n + 1, # just adding a tiny bit of jitter so as to more easily see the different lines 
                       TRUE ~ n)) %>%
  ggplot(aes(x = seasonU, y = n, col = Situation))+
  geom_line(aes(group = interaction(Situation, Vultures), linetype = Vultures), linewidth = 1)+
  geom_point(size = 4, aes(pch = Vultures))+
  scale_shape_manual(values = c(19, 1))+
  theme_classic()+
  ylab("# vultures")+
  xlab("Season")+
  scale_color_manual(values = c(cc[["feedingColor"]], cc[["flightColor"]], cc[["roostingColor"]]))+
  theme(text = element_text(size = 16))

wide <- focals %>%
  pivot_wider(id_cols = c("season", "year", "Situation", "seasonUnique", "seasonU"), names_from = "Vultures", values_from = "n")
```

Networks varied in the total and focal number of individuals, from a minimum of `r min(wide$total)` vultures (`r wide$focal[wide$total == min(wide$total)][1]` focal) in the `r wide$year[wide$total == min(wide$total)][1]` `r wide$season[wide$total == min(wide$total)][1]` season to a maximum of `r max(wide$total)` vultures (`r wide$focal[wide$total == max(wide$total)]` focal) in the `r wide$year[wide$total == max(wide$total)]` `r wide$season[wide$total == max(wide$total)]` season in the co-roosting network (@fig-nvultures) . The number of vultures tagged generally increased over time as more individuals were fitted with tags over the course of the study period.

```{r}
#| echo: false
#| output: true
#| label: fig-density
#| fig-cap: "Density of the proximity-based social networks created for each season and social situation."

load("data/derived/flightSeasons_mode10.Rda")
load("data/derived/feedingSeasons_mode10.Rda")
load("data/derived/roostSeasons.Rda")

flightSeasons_mode10_g <- map(flightSeasons_mode10, ~vultureUtils::makeGraph(mode = "sri", data = .x, weighted = T))
feedingSeasons_mode10_g <- map(feedingSeasons_mode10,  ~vultureUtils::makeGraph(mode = "sri", data = .x, weighted = T))
roostSeasons_g <- map(roostSeasons,  ~vultureUtils::makeGraph(mode = "sri", data = .x, weighted = T))

# Describing the networks -------------------------------------------------
# Network-level measures
fldens <- map(flightSeasons_mode10_g, igraph::edge_density) %>% unlist()
fedens <- map(feedingSeasons_mode10_g, igraph::edge_density) %>% unlist()
rodens <- map(roostSeasons_g, igraph::edge_density) %>% unlist()
dens <- data.frame(season = factor(seasonNames, levels = seasonNames), flight = fldens, feeding = fedens, roosting = rodens) %>%
  pivot_longer(cols = c("flight", "feeding", "roosting"), names_to = "Situation", values_to = "Density") %>%
  mutate(seasonU = str_replace(season, "breeding", "B")) %>%
  mutate(seasonU = str_replace(seasonU, "fall", "F")) %>%
  mutate(seasonU = str_replace(seasonU, "summer", "S")) %>%
  mutate(seasonU = str_remove(seasonU, "_")) %>%
  mutate(seasonU = str_remove(seasonU, "(?<=^)20")) %>%
  mutate(seasonU = factor(seasonU, levels = c("20F", "21B", "21S", "21F", "22B", "22S", "22F", "23B")))

dens %>%
  ggplot(aes(x = seasonU, y = Density, col = Situation))+
  geom_line(aes(group = Situation), linewidth = 1)+
  geom_point(size = 4)+
  theme_classic()+
  ylab("Network density")+
  xlab("Season")+
  scale_color_manual(values = c(cc[["feedingColor"]], cc[["flightColor"]], cc[["roostingColor"]]))+
  theme(text = element_text(size = 16))
```

Seasonal social networks varied in density from `r min(dens$Density)` (`r stringr::str_extract(dens$season[dens$Density == min(dens$Density)], "[a-z]+")` `r str_extract(dens$season[dens$Density == min(dens$Density)], "[0-9]+")` ) to `r max(dens$Density)` (`r stringr::str_extract(dens$season[dens$Density == max(dens$Density)], "[a-z]+")` `r str_extract(dens$season[dens$Density == max(dens$Density)], "[0-9]+")`) (@fig-density). Co-roosting networks were generally denser than co-feeding or co-flight networks.

Two-way interactions between situation and both movement (PC1) and exploration (PC2) were retained in the degree model. No interaction effects including season were retained, and the main effect of season was not significant (Table XX).

@@

\[text here\]

>>>>

```{r}
#| include: false
library(tidyverse)
library(here)
```

@@

Social situations occur over different time scales, and this is an important consideration when studying the causes and consequences of their structure. In defining co-roosting, co-flight, and co-feeding interactions based on proximity in each of these behavioral situations, we have created social networks in which edges already conceptually represent interactions over different amounts of time. Because individuals have one roosting location per night, edge weights in the co-roosting network are measured in units of days, while pairs of individuals can interact many times in flight or at a carcass on a given day.\
\
The biological processes driving the execution of these behaviors and the formation of these networks are likewise happening over different time scales. An individual must decide once per day which roost to go to [@harel2016]. But flight is a continuous process driven by many decisions resulting from individual and social information [@sassi2023], and feeding events are concentrated in time on a given day. Feeding events are driven by the temporal rhythm of the appearance of carcasses, which occur naturally or are provisioned every \[NUMBER\] days and persist for hours to a few days (Orr Spiegel, pers. comm.). Individuals may return to the same carcass on successive days. While at a carcass, individuals may fight and interact many times.

@@

To answer these questions, we analyzed data from a free-flying population of Eurasian griffon vultures (Gyps fulvus) fitted with GPS tags.

\[tag specs\]

Over the period \[date\] to \[date\], the dataset contained \[number\] of tagged vultures. Tags recorded GPS fixes, as well as speed and heading readings, every 10 minutes when battery conditions allowed. The GPS data are stored on Movebank [@kays2022] and were loaded into R using the `move` package\[ @kranstauber2023\].

@@

We subsetted the dataset to exclude periods of time that vultures spent in capture cages, in the wildlife hospital following poisonings or injuries, or in captivity prior to wild release (in the case of captive-born birds). We cleaned the data to remove invalid GPS fixes. Then we split data from each into three seasons. The fall, or "pre-breeding," season includes dates from 15 September through 14 December; the breeding season lasts from 15 December through 14 May of the following year; and the summer lasts from 15 May through 14 September. These season boundaries were chosen based on biological knowledge of the vultures' breeding timelines (see more details in the Study System section). Therefore, the analysis contained a total of 8 seasons, from the fall of 2020 through the 2023 breeding season (Table X). All of these data was reserved for use in proximity-based social network analysis. Since the GPS tags give few to no fixes overnight, and since reception is often very poor on the cliffs where vultures roost, we combined information from individuals' last locations of the evening and first locations of the morning to assign a roost site for each individual on each night \[more details\]. Full details about the data cleaning and roost assignment procedure can be found in the supplementary material.

@@

For each season, we calculated movement measures only for individuals tracked for at least 30 days (after removing low-point days as described above). We calculated flight altitude by extracting ground elevation information from elevation rasters obtained from the `elevatr` R package [@hollister2022], at a zoom level of \[level–more details about resolution?\] and subtracting those numbers from the vultures' tag-mounted altimeter readings. Finally, since some tags were occasionally set to a much higher fix rate (e.g. every 1 minute) for short periods of time, we downsampled the data to a 10 minute fix rate to make sure that all individuals had comparable fix rates.

@@

For each focal individual, we characterized a pattern-based spatial phenotype (Webber et al. 2023) by calculating a variety of movement and space use measures over the course of each of the eight seasons. We calculated home range as the 95% AKDE of a vulture's GPS fixes in a season @

@@

To summarize individuals' spatial phenotype, we created three composite indices: **movement**, **space use**, and **roost diversification**. Each index was derived from a principal component analysis of several spatial phenotype variables, all of which were scaled and centered over the entire date range of the study before putting them into the PCA.

**Movement** was defined as the first principal component of a PCA containing per-vulture seasonal means of daily mean flight altitude, daily flight duration, daily maximum displacement, and daily flight tortuosity. **Space use** was defined as the first principal component of a PCA containing log-transformed 95% AKDE home range sizes and the ratio of the 50% AKDE to the 95% AKDE. **Roost diversification** was defined as the first principal component of a PCA containing a seasonal Shannon diversity index of roost sites used \[ref Shannon\], number of unique roost sites used, and the proportion of nights in a given season when a vulture switched to a new roost site (versus roosting at the same site as on the previous night). The composite measures accounted for %, %, and % of the variation in their respective PCAs.

@@

```{r}
#| include: false
source(here("scripts/04.0_mixedModels.R"))
```

```{r}
#| echo: false
#| label: fig-d-space
#| fig-cap: "Relationship between normalized degree and space use"
plot_d_eff_space
```

```{r}
#| echo: false
#| label: fig-d-roost
#| fig-cap: "Relationship between normalized degree and roost diversification"
plot_d_eff_roost
```

```{r}
#| echo: false
#| label: fig-d-space-situ
#| fig-cap: "Relationship between normalized degree and space use, with separate trend lines for each social situation"
plot_d_eff_situ_space
plot_d_emt_situ_space
```

```{r}
#| echo: false
#| label: fig-s-movement
#| fig-cap: "Relationship between normalized strength and movement"
plot_s_eff_movement
```

```{r}
#| echo: false
#| label: fig-d-movement-situ
#| fig-cap: "Relationship between normalized degree and movement, with separate trend lines for each social situation"
plot_d_eff_situ_movement
plot_d_emt_situ_movement
```

```{r}
#| echo: false
#| label: fig-s-roost
#| fig-cap: "Relationship between normalized strength and roost diversification"
plot_s_eff_roost
```

```{r}
#| echo: false
#| label: fig-s-space
#| fig-cap: "Relationship between normalized strength and space use"
plot_s_eff_situ_space
plot_s_emt_situ_space
```

@@
